\documentclass[11pt,twoside,openright]{report}
%\usepackage[T1]{fontenc}
%\usepackage[english,french]{babel}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algcompatible}
\usepackage[noend]{algpseudocode}
\usepackage{fancyvrb}
\usepackage{hyperref}
\usepackage{subfigure}
\usepackage{graphicx}
\usepackage{blkarray}
\usepackage{pgfplots}
\usepackage{multirow}
\usepackage{siunitx}
\usepackage[framemethod=tikz]{mdframed}
\usepackage{multicol}
\usepackage{pdfpages}
\definecolor{mycolor}{rgb}{0.122, 0.435, 0.698}

\newmdenv[innerlinewidth=0.5pt, roundcorner=4pt,linecolor=mycolor,innerleftmargin=6pt, innerrightmargin=6pt,innertopmargin=6pt,innerbottommargin=6pt]{mybox}

\pgfplotsset{
    compat=1.15,
    %every axis legend/.append style={at={(0.5,-0.13)},anchor=north,legend cell align=left},
    legend pos=north west,
    legend style={draw=none,
              text width=0.9in,           %% adjust
              %minimum height=0.5in,     %% adjust
              %anchor=center,
              %cells={anchor=west},
            },
    %axis lines=left,
}

\usepackage{tikz}
\usetikzlibrary{arrows, automata,calc}
\usetikzlibrary{er,positioning}
\tikzstyle{block} = [rectangle, draw, fill=blue!20, 
    text width=6em, text centered, rounded corners, minimum height=4em]
\tikzstyle{line} = [draw, -latex']

\newcommand\irregularcircle[2]{% radius, irregularity
  \pgfextra {\pgfmathsetmacro\len{(#1)+rand*(#2)}}
  +(0:\len pt)
  \foreach \a in {10,20,...,350}{
    \pgfextra {\pgfmathsetmacro\len{(#1)+rand*(#2)}}
    -- +(\a:\len pt)
  } -- cycle
}
%\usepackage{etoolbox}
%\apptocmd{\sloppy}{\hbadness 10000\relax}{}{}
\apptocmd{\thebibliography}{\raggedright}{}{}

\input{macros-ph}
\input{macros-abstr}
\input{tikzstyles2}
\makeatletter
\def\input@path{{figures/}}
\graphicspath{{figures/}}
\makeatother
\newcommand{\acm}[3]{\{#1\}\rightarrow#3}
\newcommand{\ac}[3]{$\{#1\}\rightarrow#3$}
\newcommand{\acmph}[3]{#1\rightarrow#2\Rsh#3}
\newcommand{\acph}[3]{$#1\rightarrow#2\Rsh#3$}
\newcommand{\omesi}{^\omega_\alpha}
\newcommand{\vv}{\mathrm{v}^{val}}
\newcommand{\dv}[2]{\dfrac{{\rm d}#1}{{\rm d}#2}}
\newcommand{\ddv}[2]{\frac{{\rm d}#1}{{\rm d}#2}}
\newcommand{\st}{{\mathrm{state}}}
\newcommand{\sol}{{\mathrm{sol}}}
\newcommand{\false}{\mathbf{False}}
\newcommand{\true}{\mathbf{True}}
\DeclarePairedDelimiter{\round}{\lfloor}{\rfloor}
\DeclarePairedDelimiter{\ceiling}{\lceil}{\rceil}
\DeclarePairedDelimiter{\init}{\langle}{\rangle}
\DeclarePairedDelimiter{\addset}{\cup\{}{\}}
\DeclarePairedDelimiter{\miset}{\setminus\{}{\}}
\theoremstyle{definition}

\newtheorem{definition}{Definition}[chapter]
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{example}{Example}[chapter]
\newtheorem{proposition}{Proposition}[chapter]
\newtheorem{remark}{Remark}[chapter]
%\newtheorem{proof}{Proof}[chapter]

\algdef{SE}[DOWHILE]{Do}{doWhile}{\algorithmicdo}[1]{\algorithmicwhile\ #1}%

\usepackage{ulem}
\newcommand{\roux}[2]{{\sout{#1}}{{\color{purple}~#2}}} %\renewcommand{\roux}[2]{{#1}{\phantom{#2}}} ou %\renewcommand{\roux}[2]{{#1}{}

\hyphenation{apprenti-ssage}


%\newenvironment{poliabstract}[1]{\renewcommand{\abstractname}{#1}\begin{abstract}}{\end{abstract}}
%\begin{poliabstract}{R\'esum\'e}
%   blah blah blah...
%\end{poliabstract}

\title{Reachability Analysis and Revision of Dynamics of Biological Regulatory Networks
\\
\vspace{0.5cm}
\large Analyse d'accessibilit\'e et r\'evision de la dynamique dans les r\'eseaux de r\'egulations biologiques
}

\date{April 2019}
\author{Xinwei Chai}
\begin{document}
%\maketitle
\includepdfmerge{Thesis_cover.pdf,1,3}
\pagenumbering{roman} 
\setcounter{page}{1}




\tableofcontents
\listoffigures
\listoftables

\begin{abstract}
Concurrent systems have been of interest for decades.
With their simple but expressive semantics, concurrent systems become a good choice to fit the data and analyze the underlying mechanics.
However, \textit{learning} and \textit{analyzing} such concurrent systems are computationally difficult.
When dealing with big data sets, the state-of-the-art techniques appear to be insufficient, either in term of efficiency or in term of precision.

In this thesis, we propose a refined modeling framework ABAN (Asynchronous Binary Automata Network) and develop reachability analysis techniques based on ABAN: PermReach (Reachability \textit{via} Permutation search) and ASPReach (Reachability \textit{via} Answer Set Programming).
Then we propose two model learning/constructing methods: CRAC (Completion \textit{via} Reachability And Correlations) and M2RIT (Model Revision \textit{via} Reachability and Interpretation Transitions) using respectively continuous and discrete data to fit the model and using reachability properties to constrain the output models.

Chapter~\ref{chap:intro} states briefly the background and the contribution of our research.

Chapter~\ref{chap:stateOfTheArt} introduces the state of the art on modeling frameworks, model checkers, different update schemes of modelings and model learning techniques.
Some of them are referenced in the following chapters.

Chapter~\ref{chap:refinement} presents our modeling framework and its related reachability analyzers based on static analysis.
We focus on the inconclusive cases of pure static analysis and extract the key components preventing from a direct solution.
We then apply heuristics on these components, solving them with a limited search to reach a more conclusive result of the reachability problem.

Chapter~\ref{chap:modelInference} presents the methodology of model learning.
Our model learners CRAC and M2RIT perform in fact model selection. 
They choose a model from the candidates satisfying the provided reachability constraints.
However the number of candidate models can be exponential, our model learners can shrink the search space with constraints when generating the models.

Chapter~\ref{chap:test} shows some comparative and exploratory tests and their results on the methods presented in Chapter~\ref{chap:refinement} and Chapter~\ref{chap:modelInference}.
PermReach and ASPReach are more efficient than traditional model checkers on the reachability analysis and they perform a more conclusive analysis while holding the running time in the same scale as pure static analyzers have.

Chapter~\ref{chap:conclusion} concludes the thesis and proposes some possible future work.
\end{abstract}

\newpage
\renewcommand{\abstractname}{R\'esum\'e} 
\begin{abstract}
\pagestyle{empty}
Les syst\`emes concurrents pr\'esentent un int\'er\^et depuis des d\'ecennies.
Avec leur s\'emantique simple mais expressive, les syst\`emes concurrents deviennent un bon choix pour ajuster les donn\'ees et analyser les m\'ecanismes sous-jacents.
Cependant, \textit{l'apprentissage et l'analyse} de tels syst\`emes concurrents sont difficiles pour ce qui concerne les calculs. %\roux{en calcul}{pour ce qui concerne les calculs}.
Lorsqu'il s'agit de grands ensembles de donn\'ees, les techniques les plus r\'ecentes semblent insuffisantes, que ce soit en termes d'efficacit\'e ou de pr\'ecision.

Dans cette th\`ese, nous proposons un cadre de mod\'elisation raffin\'e ABAN (Asynchronous Binary Automata Network) et d\'eveloppons des techniques d'analyse d'atteignabilit\'e bas\'ees sur ABAN: PermReach (Reachability \textit{via} Permutation search) et ASPReach (Reachability \textit{via} Answer Set Programming).
Nous proposons ensuite deux m\'ethodes de construction et d'appren-tissage %\roux{d'apprentissage}{de construction et d'apprentissage} 
des mod\`eles: CRAC (Completion \textit{via} Reachability And Correlations) et M2RIT (Model Revision \textit{via} Reachability and Interpretation Transitions) en utilisant respectivement des donn\'ees continues et discr\`etes pour s'ajuster au mod\`ele et des propri\'et\'es d'accessibilit\'e afin de contraindre les mod\`eles r\'esultants.% \roux{pour contraindre les mod\`eles en sortie}{afin de contraindre les mod\`eles résultants}.

Le chapitre~\ref{chap:intro} d\'ecrit bri\`evement le contexte et la contribution de nos recherches.
Le chapitre~\ref{chap:stateOfTheArt} pr\'esente l'\'etat de l'art des mod\'elisations, des model checkers, des diff\'erentes dynamiques associ\'e aux mod\`es et les techniques d'apprentissage des mod\`eles.
Certains d'entre eux sont r\'ef\'erenc\'es dans les chapitres suivants.

Le chapitre~\ref{chap:refinement} pr\'esente notre cadre de mod\'elisation et ses analyseurs d'accessibilit\'e associ\'es, qui sont %\roux{li\'es}{associ\'es, qui sont}
 bas\'es sur l'analyse statique.
Nous nous concentrons sur les cas non concluants d'analyse statique pure et extrayons les composants cl\'es emp\^echant une solution directe.
Nous appliquons ensuite des heuristiques sur ces composants, en les r\'esolvant avec une recherche limit\'ee pour obtenir un r\'esultat plus concluant du probl\`eme d'accessibilit\'e.

Le chapitre~\ref{chap:modelInference} pr\'esente la m\'ethodologie de l'apprentissage par mod\`ele.
Nos syst\`emes de construction de mod\`eles par apprentissage %\roux{apprenants mod\`eles}{syst\`emes de construction de mod\`eles par apprentissage}
 CRAC et M2RIT effectuent en fait une s\'election des mod\`eles.
Ils choisissent un mod\`ele parmi les candidats qui satisfont \`a toutes les contraintes d'accessibilit\'e donn\'ees.
Cependant, le nombre de mod\`eles candidats pouvant \^etre de tr\`es grande taille, %\roux{exponentiels}{de tr\`es grande taille}
 nos r\'eviseurs de mod\`eles peuvent r\'eduire l'espace de recherche avec des contraintes lors de la g\'en\'eration des mod\`eles.

Le chapitre~\ref{chap:test} pr\'esente quelques tests comparatifs et exploratoires et leurs r\'esultats sur les m\'ethodes pr\'esent\'ees aux chapitres~\ref{chap:refinement} et \ref{chap:modelInference}.
PermReach et ASPReach sont plus efficaces que les v\'erificateurs de mod\`ele traditionnels pour l'analyse de l'accessibilit\'e. Ils effectuent une analyse plus concluante tout en maintenant la dur\'ee de fonctionnement \`a la m\^eme \'echelle que les analyseurs statiques purs.

Le chapitre~\ref{chap:conclusion} conclut la th\`ese et propose des travaux futurs possibles.
\end{abstract}

\newpage
\chapter*{List of symbols}
\thispagestyle{empty}
\begin{tabular}{l|l}
    $\land$ & logic AND\\
    $\lor$ & logic OR\\
    $\oplus$ & logic XOR\\
    $|A|$ & the cardinality of set $A$\\
    $[i,j]$& the interval comprising real numbers $x$ satisfying $i\leq x\leq j$\\
    $[i;j]$& the interval comprising integers $x$ satisfying $i\leq x\leq j$\\
    $C_n^k$& the $k$-combination in the set of $n$ elements, $C_n^k=\frac {n!}{k!(n-k)!}$\\
    $a_i$  & automaton $a$ is taking value $i$ \\
    $x::y$ & event $x$ happens just before $y$ in the sequence\\
    $a.next$&the successor of $a$\\
    $a.pred$ & the predecessor of $a$\\
    $A\to b_j$ & condition $A$ allows automaton $b$ to reach qualitative level $j$\\
    $\mathbb{N}$ & the set of all natural numbers\\
    $\bar{x}$ & the mean of variable $x$
\end{tabular}
\newpage
\pagenumbering{arabic} 
\input{Chap1-Introduction.tex}

\input{Chap2-State-Of-The-Art.tex}

\input{Chap3-Refinements-Of-Reachability-Analysis.tex}

\input{Chap4-Model-Inference-And-Revision.tex}

\input{Chap5-Tests-And-Benchmarks.tex}

\input{Chap6-Conclusion-And-Future-Work.tex}

\bibliographystyle{plain}
\bibliography{bib}

\appendix
\input{Chap7-Appendix.tex}
\includepdfmerge{Thesis_cover.pdf,2}
\end{document}